---
title: "Processing Clinical Trial Analysis Data with the {forceps} Package"
author: "Michael Kane<br><span style = 'font-size: 80%;'>School of Public Health, Biostatistics Department<br>Yale University<br>"
output:
  xaringan::moon_reader:
    css: [duke-blue, metropolis, tamu-fonts]
    titleSlideClass: [left, middle]
---

```{r echo = FALSE}
xaringanExtra::use_panelset()
```

# <br>Background: Sumbmitting study results to the FDA

<br>
## Clinical trial study data being submitted to the Food and Drug Administration 
## must conform to the standards developed by the Clinical Data Interchange 
## Standards Consortium (CDISC).

--

<br>
## The goal is to "enable information system interoperability to 
## improve medical research and related areas of healthcare."

--

<br>
## Submissions include Analytical Data Model (ADaM) formatted data
- Individual-level (patient) data
- Data values are validated by at least two analysts 
- Used directly in statistical analyses to show a therapies efficacy

---

# <br> Why might we care about ADaM formatted data?

<br>
## We might be providing an analysis for a trial

--

<br>
## This is the format for popular open-access data sources
- [Project DataSphere](https://www.projectdatasphere.org/)
- [Immport](https://www.immport.org/home)

--
  
## These data facilitate a lot of biomedical research
- Prognostic characterization of patients (based on control data)
- Subtype identification and analysis

---

# <br> So, what's the challenge? Why are you giving this talk?

<br>
## The data format is SAS-centric
- longitudinal data are separated from demographic data don't mix
- we may have longitudinal data at different times per individual
- categorical variables need a string and a integer-valued variable

--

<br>
## Data sets are not tidy
- A column may correspond to multiple variables
- A separate columns are needed to identify the variable and a variable value

---

# <br> The {forceps} package

<br>

## A minimal set of functions and data that make it easy to go from ADaM 
## formatted data to a single analyzable data set.

<br>

## Verbs
- `data_desc()`
- `contradictions()`
- `consolidate()`
- `cohort()`

<br>

## Includes toy .sas7bdat files with common ADaM file issues.

---

# <br> What do the data look like?

.panelset[

.panel[.panel-name[R Code]

```{r eval = FALSE}
library(haven)
library(purrr)
fn <- dir("adam-test")
ds <- map(fn, ~ read_sas(file.path("adam-test", .x)))
names(ds) <- gsub(".sas7bdat", "", fn)
names(ds)
head(ds[[1]])
head(ds[[2]])
head(ds[[3]])
```
]

.panel[.panel-name[Output 1]
```{r eval = TRUE, echo = FALSE}
library(haven)
library(purrr)
fn <- dir("adam-test")
ds <- map(fn, ~ read_sas(file.path("adam-test", .x)))
names(ds) <- gsub(".sas7bdat", "", fn)
names(ds)
```
]

.panel[.panel-name[Output 2]
```{r eval = TRUE, echo = FALSE}
ds[[1]]
```
]

.panel[.panel-name[Output 3]
```{r eval = TRUE, echo = FALSE}
ds[[2]]
```
]

.panel[.panel-name[Output 4]
```{r eval = TRUE, echo = FALSE}
ds[[3]]
```
]

]

---

# <br> What do we want the data to look like?

<br>
```{r echo = FALSE, message = FALSE, warning = FALSE}
x <- normalizer::pnc_ds(ds, collapse_on = "PtID", handle_contra_vals = TRUE, verbose = FALSE)
x[,1:10]
```

---

# <br>Create data descriptions 

```{r eval = FALSE, echo = TRUE}
data_desc(ds[[1]])
```

```{r eval = TRUE, echo = FALSE}
library(tibble)
tibble(var = names(ds[[1]]), 
       desc = c("Patient", "Adverse Event Duration", "Adverse Event", "Adverse Event Type", "Adverse Event Grade",
                "Adverse Event Treatment", "Adverse Event Day", "Worse Common Toxicity Criteria",
                "Sex", "Race", "Overall Survival Days", "Overal Survival Censor"))
```

```{r eval = TRUE, echo = FALSE}
contras <- normalizer::contradiction_tibbles(ds, collapse_on = "PtID")
```

---
# <br>Find conflicts

.panelset[

.panel[.panel-name[R Code]
```{r eval = FALSE, echo = TRUE}
contras <- contradictions(ds, on = "PtID")
contras
```
]

.panel[.panel-name[contras[1]]
```{r eval = TRUE, echo = FALSE}
contras[1]
```
]

.panel[.panel-name[contras[2]]
```{r eval = TRUE, echo = FALSE}
contras[2]
```
]

.panel[.panel-name[contras[3]]
```{r eval = TRUE, echo = FALSE}
contras[3]
```
]

]

---

# <br>Second, pivot_wider() on mutiple-variable columns

```{r messages = FALSE, warnings = FALSE, echo = TRUE, eval = TRUE}
ds$ae
```

---

# <br>Second, pivot_wider() on mutiple-variable columns

<br>
```{r messages = FALSE, warnings = FALSE, echo = TRUE, eval = TRUE}
library(dplyr, quietly = TRUE, warn.conflicts = FALSE)
library(tidyr, quietly = TRUE, warn.conflicts = FALSE)
library(foreach, quietly = TRUE, warn.conflicts = FALSE)
options(dplyr.summarise.inform = FALSE)

ae_rep <- ds$ae %>% 
  group_by(PtID) %>%
  summarize(PtID = PtID[1],
            Sex = Sex[1],
            Race = Race[1],
            osdays = osdays[1],
            os_censor = os_censor[1])
```

---

# <br>Second, pivot_wider() on mutiple-variable columns

<br>

```{r messages = FALSE, warnings = FALSE, echo = TRUE, eval = TRUE}
ae_wider <- foreach(it = split(seq_len(nrow(ds$ae)), ds$ae$PtID), 
                    .combine = bind_rows) %do% {
  tryCatch(ds$ae[it,] %>% 
             select(PtID, AE, Grade) %>%
             distinct(PtID, AE, .keep_all = TRUE) %>%
             pivot_wider(names_from = AE, values_from = Grade),
           error = function(e) tibble(PtID = ds$ae[it,]$PtID))
}

for (j in seq_along(ae_wider)) {
  ae_wider[[j]][is.na(ae_wider[[j]])] <- 0
}

ds$ae <- full_join(ae_rep, ae_wider, by = "PtID")
```

---

# <br> Try to consolidate() the data sets

```{r, eval = FALSE, fig.width = 6, fig.height=4}
x <- consolidate(ds, on = "PtID")
dim(x)
x[1:5, 1:10]
```

```{r, echo = FALSE, warning = FALSE, message=FALSE}
x <- normalizer::consolidate(ds, collapse_on = "PtID", verbose = FALSE)
dim(x)
x[1:5, 1:10]
```

---

# <br> The cohort() Verb

<br>
## Cohort:
- A demographic grouping of people
- To associate with such a group

<br>

## There is a `cohort()` function that nests on variable value groups
- Values that are repeated for the group are retained in the cohorted variable
- Other variables are put into a list column

<br>

## Useful for doing things like nesting on site id for example
- Write a function to pull features from the nested values and compare across sites
- Could be use to answer: do certain sites see better/worse response?
---

# <br> Can I use it now?

<br>

## Package has been used to harmonize 12 different analyses

--

<br>

--

## The package will be up in the next few weeks

<br>

--

## The interface should be considered stable

<br>

--

## It will be available at [https://github.com/kaneplusplus/forceps](https://github.com/kaneplusplus/forceps)

<br>

--

## Thanks!